{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-07 thesis/tm.tex
handle some todos in txn

% or "Contribution"
\revision{{\bf Motivation.}
For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Input nondeterminism is another research field,
addressed by symbolic execution \cite{symbolic}; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

@@

\begin{proof}
	Case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/\allowbreak{}write or write/write)
	with $\tiiat{\beta_1}\dots\tiiat{\beta_n}$.
	If so, then the hardware will abort \tii's transaction, discarding the effects of $\tiiat{\beta_1}\dots\tiiat{\beta_n}$
	and jumping to $\tiiat{\phi_1}$,
	satisfying case 1.
	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}, \cref{sec:landslide-dpor-hb}),
	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
	may be successfully executed until transaction commit,
	and reordering them produces an equivalent interleaving,
	satisfying case~2.
\end{proof}

@@

\begin{enumerate}
	\item Unit tests and microbenchmarks
	\begin{itemize}
		\item {\tt htm1}: The bug from \Cref{fig:htm-example,fig:htm-example-reproduced}.
			%(assertion failure on {\tt count}).
		\item {\tt htm2}: The fixed version as in \Cref{fig:htm-fixed}.
		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex locking failure path with an atomic {\tt xadd}, from \cite{tm-benchmark-cmu}.
		\item {\tt swap}: Microbenchmark that swaps values in an array, from \cite{tm-benchmark-cmu}.
		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
			% TODO: make sure figure number reference is right
		\item {\tt fig63}: Generalized version of \Cref{fig:retry-set-example},
			contrived to induce as much reduction as possible from the retry sets optimization.
	\end{itemize}
	\item Data structure tests
	\begin{itemize}
		\item {\tt avl\_insert}: AVL tree concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed (spoilers!!).
		\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
		\item {\tt avl\_mutex}: {\tt avl\_fixed} with transactional sections simplified by abstraction into a mutex.
		\item {\tt map\_mutex}: {\tt map\_basic} simplified likewise.
		\item {\tt map\_basicer\_mutex}: {\tt map\_basicer} simplified likewise.
	\end{itemize}
	\item Lock abstraction tests
	\begin{itemize}
		\item {\tt lock}(): Checks that multiple threads using a transactional lock
			cannot access the critical section simultaneously.
			% htm_spinlock and htm_mutex
		\item {\tt lock\_fast}(): Checks that a transactional lock's fast path will not suffer conflict aborts
			if its client threads' critical sections are independent.
			% htm_spinlock_fastpath and htm_mutex_fastpath
	\end{itemize}
		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
		{\tt spin\_\allowbreak{}fixed} (spoilers!!),
		and {\tt mutex} (replaces the spinlock with a Landslide-annotated P2 mutex to reduce state space size).
\end{enumerate}

@@

			% TODO: check these again, will 2,4 even complete on mutex
			{\tt lock}({\tt spinlock})	& 1	& 2	& 4	& 28	\\ % 2,2
			{\tt lock}({\tt spin\_fixed})	& 1	& 2	& 4	& 28	\\ % 2,2 or 4,1
			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 30	\\ % 2,3 gives 30; 2,4 would give 40
			% not putting lock_fast because it like, doesn't have a "state space" per se
		\end{tabular}
	\end{center}
	\caption[Number of concurrency events per iteration of each test case.]
		{Number of concurrency events per iteration of each test case.
	Note that no test used any synchronization besides mutexes % ;
	%while they do use the P2 thread API, %in the first place,
	(the P2 thread API was annotated as trusted (\cref{sec:landslide-config-landslide})
	and so does not contribute to state space size).
	Also note that ``\#race'' means the number of unique accesses identified as racy,
	rather than racing pairs
	(the other half of a pair might well be within a transaction, which cannot be preempted on).
	}
	\label{tab:tm-test-case-evence}
\end{table}

>>>>

% or "Contribution"
\revision{{\bf Motivation.}
For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Symbolic execution \cite{symbolic}, a neighbouring research area,
addresses input nondeterminism; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

@@

\begin{proof}
	Case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/\allowbreak{}write or write/write)
	with $\tiiat{\beta_1}\dots\tiiat{\beta_n}$.
	If so, then the hardware will abort \tii's transaction, discarding the effects of $\tiiat{\beta_1}\dots\tiiat{\beta_n}$
	and jumping to $\tiiat{\phi_1}$,
	satisfying case 1.
	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}; see also \cref{sec:landslide-dpor-hb}),
	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
	may be successfully executed until transaction commit,
	and reordering them produces an equivalent interleaving,
	satisfying case~2.
\end{proof}

@@

\begin{enumerate}
	\item Unit tests and microbenchmarks
	\begin{itemize}
		\item {\tt htm1}: The bug from \Cref{fig:htm-example,fig:htm-example-reproduced}.
			%(assertion failure on {\tt count}).
		\item {\tt htm2}: The fixed version as in \Cref{fig:htm-fixed}.
		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex locking failure path with an atomic {\tt xadd}, from \cite{tm-benchmark-cmu}.
		\item {\tt swap}: Microbenchmark that swaps values in an array, from \cite{tm-benchmark-cmu}.
		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
		\item {\tt fig63}: Generalized version of \Cref{fig:retry-set-example},
			contrived to induce as much reduction as possible from the retry sets optimization.
	\end{itemize}
	\item Data structure tests
	\begin{itemize}
		\item {\tt avl\_insert}: AVL tree concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed (spoilers!!).
		\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
		\item {\tt avl\_mutex}: {\tt avl\_fixed} with transactional sections simplified by abstraction into a mutex.
		\item {\tt map\_mutex}: {\tt map\_basic} simplified likewise.
		\item {\tt map\_basicer\_mutex}: {\tt map\_basicer} simplified likewise.
	\end{itemize}
	\item Lock abstraction tests
	\begin{itemize}
		\item {\tt lock}(): Checks that multiple threads using a transactional lock
			cannot access the critical section simultaneously.
			% htm_spinlock and htm_mutex
		\item {\tt lock\_fast}(): Checks that a transactional lock's fast path will not suffer conflict aborts
			if its client threads' critical sections are independent.
			% htm_spinlock_fastpath and htm_mutex_fastpath
	\end{itemize}
		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
		{\tt spin\_\allowbreak{}fixed} (spoilers!!),
		and {\tt mutex} (replaces the spinlock with a Landslide-annotated P2 mutex to reduce state space size).
\end{enumerate}

@@

			{\tt lock}({\tt spinlock})	& 1	& 2	& 4	& 28	\\ % 2,2
			{\tt lock}({\tt spin\_fixed})	& 1	& 2	& 4	& 28	\\ % 2,2 or 4,1
			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 40	\\ % 2,4
			% not putting lock_fast because it like, doesn't have a "state space" per se
		\end{tabular}
	\end{center}
	\caption[Number of concurrency events per iteration of each test case.]
		{Number of concurrency events per iteration of each test case.
	Note that no test used any synchronization besides mutexes % ;
	%while they do use the P2 thread API, %in the first place,
	(the P2 thread API was annotated as trusted (\cref{sec:landslide-config-landslide})
	and so does not contribute to state space size).
	Also note that ``\#race'' means the number of unique accesses identified as racy,
	rather than racing pairs
	(the other half of a pair might well be within a transaction, which cannot be preempted on).
	}
	\label{tab:tm-test-case-evence}
\end{table}
